{RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-03 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Finished polishing.

@@

\subsection{Comparison to other methods}
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to 13 other competitors from the literature. The numerical results are reported in Tables~\ref{tab:comp_noise0}-\ref{tab:comp_noise001}. The results for \textsc{Feynman} and \textsc{Storgatz} are given w.r.t.\  the following levels of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise). More precisely, the white Gaussian noise is added to the target variable the same way as in~\cite{la2021contemporary}:
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

@@

\begin{itemize}
	\item When considering the instance problems with no-noise, the lowest average runtime is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228 sec.). A slightly worse runtime is obtained by \textsc{DSR} algorithm (232 sec.). Note that the relative exact percentage produced by \textsc{RILS}-\textsc{ROLS} is far better from \textsc{DSR} (60.62\% versu 19.71\%). All remaining approaches deliver an order of magnitude larger average runtimes than that of \textsc{RILS}-\textsc{ROLS} and \textsc{DSR} methods..   
	\item  When considering instances with the noise level of 0.001, the lowest average runtime to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108 sec.). The second best is produced by \textsc{DSR}, but this time significantly larger (622 sec.) compared to \textsc{RILS-ROLS}. 
	\item When considering the instance problems with the noise level of 0.01, the conclusion is similar to previous. 
\end{itemize}

@@

First, the omnibus rank-based Friedman’s test for all approaches was conducted. 
The number of ranks is 1300, i.e., for each run of each method, the separate rank is calculated. Rank is based on the value of indicator variable, which takes 1 if the method delivers correct model in a given run, otherwise it takes value of 0, no matter if method finished and delivered incorrect model, or method did not finish at all.   Further, the ranks are averaged and used to test the null hypothesis, which states that there is no statistical differences between the average ranks of all competitors. 
In the case in which the null hypothesis $H_0$ was rejected, pairwise comparisons were further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. In each CD plot, the 15 competitors are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn. This analysis is done for all three levels of noise, and the corresponding CD plots are shown in Figures~\ref{fig:CDplots-no-noise}-\ref{fig:CDplots-noise0.01}.

@@

In this section we study the scalability of \textsc{RILS-ROLS} w.r.t. size of instances problems under different levels of noise.  For the sake of having simpler presentation, we split the instances from \textsc{Random} benchmark into three parts as follows. 
\begin{itemize}
	\item \textit{Small-sized instances}: instances with solution size from 3 to 7.
	\item \textit{Medium-sized instances}: those with size from 8 to 11.
	\item \textit{Large-sized instances}: instances with size from 12 to 15. 
\end{itemize}

@@

\fxnote{ACA: do ovde stigao}

When one concerns of the runtime analysis of \textsc{RILS}--\textsc{ROLS} algorithm on the \textsc{Random} benchmark set w.r.t\ number of variables, the results are displayed in Figure~\ref{fig:runtime_rils_rols}. The instances are grouped w.r.t.\ number of variables involved into corresponding exact models ($x$-axis). The following conclusions may be drawn from there.

\begin{itemize}
	\item As one could expect, average runtime is getting increased with the increase of the number of variables in exact models. 
	\item Note that the average runtime in case of the instances whose exact models have the largest number of variables (5) is just about 200$s$.
	\item The two above-mentioned points indicate efficacy of our method especially in case of the instances with small-to-medium sized exact models in the presence of small--to--middle noise levels included in the input data. In these cases, the algorithm is able to deliver a reasonable exact percentage rate, which are always larger than 50\%. Note that models which have more compact representations (thus, small in their size) are desirable in many areas of natural sciences as they are easy interpret-able than models represented by a more complex mathematical formula.   
\end{itemize}

\begin{figure}[!ht]
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[  width=180pt,height=140pt]{plots/numsize_vs_time_no_noise.eps}
		\caption{No noise}
		\label{fig:time-no-noise}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}

@@

In this paper, we dealt with solving the well--known Symbolic regression (SR)  problem. SR problem is a generalization of more interdisciplinary known problems such as linear and logistic regression.    
We proposed a meta-heuristic approach, called \textsc{RILS}-\textsc{ROLS}, which is built upon the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the ordinary least square method to efficiently determine best-fitting internal coefficients to the candidate solution. % Thus, the search process is primarily focused on considering already promising solutions in the continuous part of the search space rather than applying enhanced heuristics to search for those coefficients. 
Another key aspect is the utilization of a carefully constructed fitness function in the search, which appeared as a combination of three important scores: RMSE, $R^2$  scores and the size of the model (solution).  Last but not least, a carefully constructed local search is applied systematically exploring solutions obtained as 1--perturbations of the considered solution, represented as trees.  1--perturbations of a solution are based on applying the six per-node change operations on the solution's tree representation.

\textsc{RILS-ROLS} method is compared to the 14 other competitor algorithms from the literature on the two ground--truth benchmark sets from the literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed the efficiency of our method which produced the best average ranking results in terms of exact percentage when compared to all other competitors on the two ground-truth benchmarks in the presence of all three different levels of noise in the input data. In all cases, new best exact percentages are obtained by our \textsc{RILS-ROLS} method.   More in details,
our algorithm was able to obtain the right model on 60.62\%, 42.08\%, and 34.77\% runs over all considered problem instances in the presence of different noise levels: 0.0, 0.001, and 0.01, respectively. For the shake of comparisons, the second best approach \textsc{AI-Feynman}, when the noise level of 0.0 and 0.001 is included, obtains the exact percentages of 52.65\%, and 31.89\%, respectively. In case of the highest noise level, the second best approach was \textsc{Afp-Fe}, obtaining an exact percentage rate of 20\%. Additionally, the scalability and robustness of \textsc{RILS}-\textsc{ROLS} method are checked and confirmed on the unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of the model, the number of variables included in exact models, and obtained average runtimes.

In future work, one could think of constructing a hybrid of \textsc{RILS-ROLS} method with some other meta-heuristics to further boost the quality of the obtained results. For example, replacing ILS scheme of \textsc{RILS-RILS} with a more general Variable neighbourhood search (VNS) scheme is a reasonable option to give it a try. VNS   utilizes a more stronger and systematic system of solutions diversification and thus escaping from local optima could be much easier and the convergence much faster.  Additionally, as proof of the efficiency and usefulness of our algorithm, this tool could be applied to solving problems from various problem domains out of the area of computer science, for example from physics or chemistry, to help for setting up reasonable hypotheses or to obtain valuable insights on obtained experimental evaluations.

\section*{Appendix -- \textsc{RILS}-\textsc{ROLS} Python package}\label{sec:appendix-1}

>>>>

Training/test data splits were randomly performed in ratio 75\%:25\%, same as in \texttt{SRBench}. Moreover, the splits are structurally the same as in \texttt{SRBench}, because we used the same set of random generator seeds and the same Python splitting method (\texttt{train\_test\_split} from Scikit-Learn library~\cite{scikit-learn}).

@@

\subsection{Comparison to other methods}
% ACA: ima 14 uporednih algoritama, ne 13
In this section, we evaluate our \textsc{RILS-ROLS} algorithm and compare it to 14 other competitors from the literature. The numerical results are reported in Tables~\ref{tab:comp_noise0}-\ref{tab:comp_noise001}. The results for \textsc{Feynman} and \textsc{Storgatz} are given w.r.t.\  the following levels of noise: 0.0 (no-noisy data), 0.001 (low-level noise), and 0.01 (high-level noise). More precisely, the white Gaussian noise is added to the target variable the same way as in~\cite{la2021contemporary}:
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{n} \sum _{i=1}^n{y_i^2}}\right),$$
where $n$ relates to the number of items in the input data.

@@

\begin{itemize}
	\item When considering the instance problems with no-noise, the lowest average runtime is reported for \textsc{RILS}-\textsc{ROLS} algorithm (228 seconds). A slightly worse runtime is obtained by \textsc{DSR} algorithm (232 seconds). Note that the relative exact percentage produced by \textsc{RILS}-\textsc{ROLS} is far better from \textsc{DSR} (60.62\% versu 19.71\%). All remaining approaches deliver an order of magnitude larger average runtimes than that of \textsc{RILS}-\textsc{ROLS} and \textsc{DSR} methods..   
	\item  When considering instances with the noise level of 0.001, the lowest average runtime to find the exact model is again delivered by \textsc{RILS}-\textsc{ROLS} algorithm (108 seconds). The second best is produced by \textsc{DSR}, but this time significantly larger (622 seconds) compared to \textsc{RILS-ROLS}. 
	\item When considering the instance problems with the noise level of 0.01, the conclusion is similar to previous. 
\end{itemize}

@@

First, the omnibus rank-based Friedman’s test for all approaches was conducted. 
The number of ranks is 1300, i.e., for each run of each method, the separate rank is calculated. Rank is based on the value of indicator variable, which takes 1 if the method delivers correct model in a given run, otherwise it takes value of 0, no matter if method finished and delivered incorrect model, or method did not finish at all.   Further, the ranks are averaged and used to test the null hypothesis, which states that there is no statistical differences between the average ranks of all competitors. 
In the case in which the null hypothesis $H_0$ was rejected, pairwise comparisons were further performed by using the Nemenyi post-hoc test~\cite{pohlert2014pairwise}. The outcome is represented by means of critical difference (CD) plots. 
% ACA: ima 15 ukupno algoritama, ne 14
In each CD plot, the 15 competitors are placed on the horizontal axis according to their average ranking. Thereafter, the CD score is computed for a significance level of 0.05. If the difference is small enough, meaning that no statistical difference is detected, a horizontal bar linking statistically equal approaches is drawn. This analysis is done for all three levels of noise, and the corresponding CD plots are shown in Figures~\ref{fig:CDplots-no-noise}-\ref{fig:CDplots-noise0.01}.

@@

In this section we study the scalability of \textsc{RILS-ROLS} w.r.t. size of instances problems under different levels of noise.
Since we do not compare \textsc{RILS-ROLS} to other competitors here, but just assess the method scalability and sensitivity to noise, the experimental setup is different. First, we run each instance only once -- note that are 5 randomly generated instances for each of 47 combinations of (size, variable count). Second, the exit criteria are different: 1) maximal running time is 300 seconds instead of 1 hour, and 2) maximal number of fitness evaluations is 100 thousands instead of 1 million. 
As before, test set takes 25\% of total data. 
For the sake of having simpler presentation, we split the instances from \textsc{Random} benchmark into three parts as follows. 
\begin{itemize}
	\item \textit{Small-sized instances}: instances with solution size from 3 to 7.
	\item \textit{Medium-sized instances}: those with size from 8 to 11.
	\item \textit{Large-sized instances}: instances with size from 12 to 15. 
\end{itemize}

@@

\textsc{RILS}-\textsc{ROLS} running times for \textsc{Random} benchmark are shown in Figure~\ref{fig:runtime_rils_rols}. Instances are grouped w.r.t.\ number of variables ($x$-axis). 
As one could expect, average running time is getting increased with the increase of the number of variables. 
\textsc{RILS-ROLS} usually terminates because the maximal number of fitness evaluation is reached (100 thousands), therefore average running times are bellow 300 seconds, which is time-based exit criterion.

\begin{figure}[!h]
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[  width=180pt,height=140pt]{plots/numsize_vs_time_no_noise.eps}
		\caption{No noise}
		\label{fig:time-no-noise}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}

@@

In this paper, we dealt with solving the well-known symbolic regression (SR)  problem.  
We proposed a meta-heuristic approach, called \textsc{RILS}-\textsc{ROLS}, built upon the iterated local search (ILS) scheme. The crucial element of this scheme is the ordinary least square method (OLS), used to efficiently determine best-fitting linear coefficients within solution equations. 
Another key aspect is the utilization of a carefully constructed fitness function. It combines three important measures of the solution: RMSE, $R^2$, and the solution size. 
Further, we designed an efficient local search that systematically explores the solution space around candidate solution.

\textsc{RILS-ROLS} method is compared to the 14 other competing methods from the literature on two ground-truth benchmark sets from the literature, named \textsc{Feynman} and \textsc{Strogatz}. The experimental evaluation confirmed the quality of our method -- \textsc{RILS-ROLS} produced the best average ranking results in terms of exact model accuracy under varying levels of noise in data. 
In details, our algorithm was able to find the right model in 60.62\%, 42.08\%, and 34.77\% under no-noise, low-level noise and high-level noise, respectively. The second best approach \textsc{AI-Feynman} obtains fairly worse percentages: 52.65\%, 31.89\% and 20\%.

In addition to well-known benchmarks from literature, we introduced a new non-biased benchmark set of randomly generated instances, labeled by \textsc{Random}. 
This benchmark was used to test \textsc{RILS}-\textsc{ROLS} scalability, by varying the size of the model and the number of variables.

In the future work, one could think of constructing a hybrid of \textsc{RILS-ROLS} method with some other meta-heuristics to further boost the quality of the obtained results. For example, replacing ILS scheme of \textsc{RILS-RILS} with a more general Variable neighbourhood search (VNS) scheme is a reasonable option. Also, \textsc{RILS-ROLS} could be used to solve practical problems from various domains. For example, it can be used in physics or chemistry to help setting-up reasonable hypotheses or to obtain valuable insights on obtained experimental evaluations. One more improvement direction is incorporating top-level constraints into the method. For example, if one knows that the modeled function needs to be monotonically decreasing or differentiable, or both, this knowledge could be used to alter the search -- eliminate inadequate parts of the search space (hard constraint) or more \emph{subtly} penalize this behavior through fitness function (soft constraint).

\fxnote{ACA: PROVERITI REFERENCE, mislim da sam negde video da nedostaju volume ili stranice ili slicno...}

\section*{Appendix - \textsc{RILS}-\textsc{ROLS} Python package}\label{sec:appendix-1}
